Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-02-12 manuscript_Huang_rev.tex
Reik's changes.

In the context of finite sample size effects, we note that for the particular example of the logistic map, one could easily generate realizations of much longer lengths. Our numerical results show, however, that longer realizations of a length of up to $L\leq5\cdot 10^6$ do not change the obtained entropy and SCM \emph{estimates} significantly (not shown). From a practical perspective, two aspects call for attention in interpreting the obtained entropy and SCM values quantitatively. On the one hand, due to the finite numerical precision of the computed trajectories, any chaotic regime may in fact turn out to appear as periodic (with arbitrarily long periods); this can lead to numerical artifacts that will be further discussed in Sec.~\ref{sec:transi}. On the other hand, there may exist asymptotic analytical results for the case of $L\to\infty$ and $D\to\infty$, e.g., for the case of $r=4$ where one would expect $\mathcal{H}_O\to 1$ and $\mathcal{C}_O\to 0$, which substantially differ from our numerical results. In this regard, all estimated values of the different entropy and complexity measures always need to be considered as finite sample estimates that may well deviate from their expected asymptotic values and thereby depend on the specific values of $L$ and $D$ available for the analysis. This is, however, a feature that is also common to other existing symbolic dynamics based entropies and SCMs.

@@

\textcolor{red}{As a first general observation}, we recover the known result that \textcolor{red}{the established} pattern frequency based SCM values \textcolor{red}{for the logistic map} lie close to the \textcolor{red}{theoretical} maximum $\mathcal{C}_{max}$ \textcolor{red}{for a given entropy value} \textcolor{red}{(Fig.~\ref{fig:CElogistic}(a, d))}, while \textcolor{red}{some residual variation of the} displacement of the estimated values from \textcolor{red}{those} upper bounds \textcolor{red}{is retained for all considered values of} the embedding dimension $D$. Most notably, we find many combinations of complexity--entropy values in the lower \emph{right} part of the plane (high entropy, low complexity), which \textcolor{red}{is commonly indicative of stochastic processes and has been visually} highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values \textcolor{red}{(see also our previous comment on the expected asymptotic value for $r=4$)}. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information \textcolor{red}{into account} in the sense that only pattern frequencies have been considered in their definitions. \textcolor{red}{The observed behavior confirms previous findings that distinguishing (especially high-dimensional) chaotic from stochastic dynamics can be a challenging task in established ordinal pattern based time series analysis methods~\cite{AmigoPTRSA2014,Riedl2013}.}

>>>>

In the context of finite sample size effects, we note that for the particular example of the logistic map, one could easily generate realizations of much longer lengths. Our numerical results show, however, that longer realizations of a length of up to $L\leq5\cdot 10^6$ do not change the obtained entropy and SCM \emph{estimates} significantly (not shown). From a practical perspective, two aspects call for attention in interpreting the obtained entropy and SCM values quantitatively. On the one hand, due to the finite numerical precision of the computed trajectories, any chaotic regime may in fact turn out to appear as periodic (with arbitrarily long periods); this can lead to numerical artifacts that will be further discussed in Sec.~\ref{sec:transi}. On the other hand, 
%there may exist asymptotic analytical results for the case of $L\to\infty$ and $D\to\infty$, e.g., for the case of $r=4$ where one would expect $\mathcal{H}_O\to 1$ and $\mathcal{C}_O\to 0$, which substantially differ from our numerical results. In this regard, 
all estimated values of the different entropy and complexity measures always need to be considered as finite sample estimates that may well deviate from their expected asymptotic values and thereby depend on the specific values of $L$ and $D$ available for the analysis. This is, however, a feature that is also common to other existing symbolic dynamics based entropies and SCMs.

@@

\textcolor{red}{As a first general observation}, we recover the known result that \textcolor{red}{the established} pattern frequency based SCM values \textcolor{red}{for the logistic map} lie close to the \textcolor{red}{theoretical} maximum $\mathcal{C}_{max}$ \textcolor{red}{for a given entropy value} \textcolor{red}{(Fig.~\ref{fig:CElogistic}(a, d))}, while \textcolor{red}{some residual variation of the} displacement of the estimated values from \textcolor{red}{those} upper bounds \textcolor{red}{is retained for all considered values of} the embedding dimension $D$. Most notably, we find many combinations of complexity--entropy values in the lower \emph{right} part of the plane (high entropy, low complexity), which \textcolor{red}{is commonly indicative of stochastic processes and has been visually} highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information \textcolor{red}{into account} in the sense that only pattern frequencies have been considered in their definitions. \textcolor{red}{The observed behavior confirms previous findings that distinguishing (especially high-dimensional) chaotic from stochastic dynamics can be a challenging task in established ordinal pattern based time series analysis methods~\cite{AmigoPTRSA2014,Riedl2013}.}
